\textls[-0]{\thetitle}

### 2023-03-23 canonicitymodel.tex
edits

\noindent
Canonicity is a fundamental property that should be satisfied by a dependent type theory serving as
a programming language or as a computational foundation for mathematics.
%\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
% We can even argue that, if this theorem is proven in a computable meta\-logic, 
% then by the Curry–Howard correspondence, the canonicity theorem provides a big-step
% interpreter for closed terms of the boolean type.
%\YZ{Can we say something about
%strong/weak normalization?}\EDJreply{
%  No. \\
%  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
%  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
%      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
%    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
%  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
%  4. Of course we can conjecture we have that normalization function \\
%  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
%  but we don't have termination (a runnable interpreter) for open terms.  \\ 
%  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
%}

@@

We state the fundamental property of logical relation here---in our proof relevant model, it is admitting that our model is really a model, and our syntax is the initial model so there is map from our syntax to our canonicity model:
\begin{lemma}[Fundamental Theorem of Logical Relation]
  If we have a (intrinsically well-typed) term $\goodTerm{\Gm}{t}{T}$, then we have its interpretation as a dependent function $\denotesC{t} : \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\denotesC{\goodType{\Gm}{T}{}}(\gm)(\gm')(\sub{t}{\gm})$
\end{lemma}

The first consequence of this model is the consistency of \TT---we cannot derive
$\goodTerm{\cdot}{t}{\bot}$. Otherwise, we would have an element in the empty set,
$\denotesC{\goodTerm{\cdot}{t}{\bot}}(\SubstEmp)(\star) \in
\denotesC{\bot}(\SubstEmp)(\star)(\sub{t}{\gm}) = \emptyset$, a contradiction.
\YZ{Have you stated the fundamental property of the logical-relations model?}\EDJreply{No. Now stated.}

@@

Further, with the help of eta rules, we have the following canonical forms
\begin{theorem}[Canonical Forms]\hfill
  \begin{itemize}
    \item If $\goodTerm{\cdot}{t}{\El{\wcode{\tau}}}$ and $\goodWSig{\cdot}{\tau}{n}$, then $\goodTerm{\cdot}{t \equiv \Wsup{j}{\tau}{a}{b}}{\El{\wcode{\tau}}}$ for some $\goodTerm{\cdot}{a}{A}$, $\goodTerm{B[("id", a)]}{b}{\El{\wcode{\tau}}}$, and $j < n$
    \item If $\goodTerm{\cdot}{t}{\mathbb{B}}$ then $\goodTerm{\cdot}{t \equiv "tt"}{\mathbb{B}}$ or $\goodTerm{\cdot}{t \equiv "ff"}{\mathbb{B}}$ 
    \item If $\goodTerm{\cdot}{t}{\TyLkg{\lsig}}$ with $\goodSig{\cdot}{\sigma}{n}$, then $\goodTerm{\cdot}{t \equiv \LkgAdd{o}{t}}{\TyLkg{\lsig}}$ 
      for some $\goodTerm{\cdot}{o}{\TyLkg{\lkgproj{1}{\sigma}}}$ and $\goodTerm{\lsigproj{2}{\sigma}}{t}{\lkgproj{2}{\sigma}}$
    \item If $\goodTerm{\cdot}{t}{\TySigma A B}$ then $\goodTerm{\cdot}{t \equiv (a, b)}{\TySigma A B}$ with $\goodTerm{\cdot}{a}{A}$ and $\goodTerm{\cdot}{b}{B[("id", a)]}$
    \footnote{We emphasize the last one because $\Tkg{\ell}$ is a dependent pair}
  \end{itemize}
\end{theorem}

>>>>

\noindent
Canonicity is a key criterion for a dependent type theory to be considered as
a programming language or as a computational foundation for mathematics.
%\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
% We can even argue that, if this theorem is proven in a computable meta\-logic, 
% then by the Curry–Howard correspondence, the canonicity theorem provides a big-step
% interpreter for closed terms of the boolean type.
%\YZ{Can we say something about
%strong/weak normalization?}\EDJreply{
%  No. \\
%  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
%  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
%      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
%    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
%  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
%  4. Of course we can conjecture we have that normalization function \\
%  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
%  but we don't have termination (a runnable interpreter) for open terms.  \\ 
%  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
%}

@@

We state the fundamental property of the logical-relations model.
%in our proof relevant model, it is admitting that our model is really a model,
%and our syntax is the initial model so there is map from our syntax to our
%canonicity model:
\begin{theorem}[Fundamental Property]
  If\/ $\goodTerm{\Gm}{t}{T}$, then its semantic interpretation is a dependent function such that
  $\denotesC{t} : \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\denotesC{\goodType{\Gm}{T}{}}(\gm)(\gm')(\sub{t}{\gm})$.
\end{theorem}

The first consequence of this model is the consistency of \TT---we cannot derive
$\goodTerm{\cdot}{t}{\bot}$. Otherwise, we would have an element in the empty set,
$\denotesC{\goodTerm{\cdot}{t}{\bot}}(\SubstEmp)(\star) \in
\denotesC{\bot}(\SubstEmp)(\star)(\sub{t}{\gm}) = \emptyset$, a contradiction.

@@

Further, with the help of eta rules, we have the following canonical forms.
\begin{theorem}[Canonical Forms]\hfill
  \begin{itemize}
    \item If $\goodTerm{\cdot}{t}{\El{\wcode{\tau}}}$ and $\goodWSig{\cdot}{\tau}{n}$, then $\goodTerm{\cdot}{t \equiv \Wsup{j}{\tau}{a}{b}}{\El{\wcode{\tau}}}$ for some $\goodTerm{\cdot}{a}{A}$, $\goodTerm{B[("id", a)]}{b}{\El{\wcode{\tau}}}$, and $j < n$
    \item If $\goodTerm{\cdot}{t}{\mathbb{B}}$ then $\goodTerm{\cdot}{t \equiv "tt"}{\mathbb{B}}$ or $\goodTerm{\cdot}{t \equiv "ff"}{\mathbb{B}}$ 
    \item If $\goodTerm{\cdot}{t}{\TyLkg{\lsig}}$ with $\goodSig{\cdot}{\sigma}{n}$, then $\goodTerm{\cdot}{t \equiv \LkgAdd{o}{t}}{\TyLkg{\lsig}}$ 
      for some $\goodTerm{\cdot}{o}{\TyLkg{\lkgproj{1}{\sigma}}}$ and $\goodTerm{\lsigproj{2}{\sigma}}{t}{\lkgproj{2}{\sigma}}$
    \item If $\goodTerm{\cdot}{t}{\TySigma A B}$ then $\goodTerm{\cdot}{t \equiv (a, b)}{\TySigma A B}$ with $\goodTerm{\cdot}{a}{A}$ and $\goodTerm{\cdot}{b}{B[("id", a)]}$
    \footnote{We emphasize the last one because $\Tkg{\ell}$ is a dependent pair}
  \end{itemize}
\end{theorem}

### 2023-03-23 metatheory-condensed.tex
edits

\item 
The next four fields model the four constructors of~\lsti{tm}.
Constructor \lsti{tm_unit} is modeled as $\Wsup{}{\wsig_{\texttt{tm}}}{\top}{x.\bot}$
and typed with $\El{\mathit{self}.\texttt{tm}}$, where $\mathit{self}$ stands for
the typing context containing the first field~\lsti{tm}.
The W-type signature $\wsig_{\texttt{tm}}$ is exposed in
this typing context; that is, \lsti{tm} has type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$.
So $\El{\mathit{self}.\texttt{tm}}$ and $\El{\wcode{\wsig_{\texttt{tm}}}}$ can be equated,
as required by rule \ruleref{tm/wsup}.

>>>>

\item 
The next four fields model the four constructors of~\lsti{tm}.
Constructor \lsti{tm_unit} is modeled as $\Wsup{}{\wsig_{\texttt{tm}}}{\top}{x.\bot}$
and typed with $\El{\mathit{self}."tm"}$, where $\mathit{self}$ stands for
the typing context containing the first field~\lsti{tm}.
The W-type signature $\wsig_{\texttt{tm}}$ is exposed in
this typing context; that is, $\mathit{self}."tm"$ has type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$.
So $\El{\mathit{self}."tm"}$ and $\El{\wcode{\wsig_{\texttt{tm}}}}$ can be equated,
as required by rule \ruleref{tm/wsup}.
